---
title: ' Spatial summation of individual cones in human color vision'
author: '[Brian P. Schmidt](https://bps10.github.io), Alexandra E. Boehm, William S. Tuten, Austin Roorda'
date: "`r Sys.Date()`"
output:
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---


```{r results="hide", message=FALSE}
setwd("~/R/TwoConeColorSummation")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(corrplot)
library(car)

# set the colors that will be used in plots below.
red <- "#e84040"
green <-"#40c631"
blue <- "#4042e8"
yellow <- "#e8df40"
gray <- "#595959"
```

```{r, message=FALSE}
trial_data <- read_csv('filtered_data.csv')

head(trial_data)
```

# Results

## Example session

```{r, fig.height=3, fig.width=4.5, message=FALSE}

(example <- trial_data %>%
  filter(., subject == '20076R', session== 4) %>%
  group_by(., masterID1, masterID2, isPair) %>% 
  summarise(., yb_mean=signif(mean(yb), 3), 
            yb_sem=signif(sd(yb) / sqrt(length(yb)), 3), 
            gr_mean=signif(mean(gr), 3), 
            gr_sem=signif(sd(gr) / sqrt(length(gr)), 3), N=length(gr),
            typeIDstr=typeIDstr[1])
)


boundary <- data.frame(x=c(-1, 0, 1, 0, -1), y=c(0, 1, 0, -1, 0))

(examplePlot <- ggplot(data=example, aes(x=yb_mean, y=gr_mean,
                                         color=interaction(typeIDstr, isPair), 
                                         shape=isPair)) +
    geom_point(size=3) + 
    geom_errorbar(aes(x=yb_mean, ymin=gr_mean - gr_sem, ymax=gr_mean + gr_sem)) +
    geom_errorbarh(aes(y=gr_mean, xmin=yb_mean - yb_sem, xmax=yb_mean + yb_sem, height=0.01)) + 
    coord_equal(xlim=c(-1, 1), ylim=c(-1, 1)) + 
    xlab('(yellow - blue) / total') + 
    ylab('(green - red) / total') + 
    scale_color_manual(
      values = c(green, red, yellow, red), 
      name="spectral type") +
    geom_text(label=c('1', '1+2', '1+3', '2', '2+3', '3'),
              nudge_x=-0.15) +
    geom_path(data = boundary, aes(x=x, y=y), colour='gray', linetype=2, inherit.aes = F) +
    #annotate("text", x = -1, y = 1, label = "B") +
    theme_bw(base_size = 15) + 
    theme(legend.position="none")
  )

ggsave("figures/example_session.pdf")


```


## Variability in sensations from cones with the same sensitivity

We first grouped each trial based on the subject and sub-type of cone or pair probed. We then found them mean, standard deviation and count for each group. The results are reported in a table below. In both subjects with classified cones, individual and pairs of M-cones produced gr means that were greater than zero, while L-cone conditions led to negative values. This observation is consistent with a predictive relationship between cone type and color report, as found previously (Sabesan et al. 2016; Schmidt et al. 2018).

```{r, fig.height=3, fig.width=4.5, message=FALSE}
(typeGroup <- trial_data %>%
  #filter(., subject != '20075L') %>% 
   group_by(., subject, masterID1, masterID2, isPair, typeIDstr) %>%
   summarise(., yb_meanC=signif(mean(yb), 3), 
             yb_semC=signif(sd(yb) / sqrt(length(yb)), 3), 
             gr_meanC=signif(mean(gr), 3), 
             gr_semC=signif(sd(gr) / sqrt(length(gr)), 3), N=length(gr)) %>%
   group_by(., typeIDstr, isPair) %>%
   summarise(., yb_mean=signif(mean(yb_meanC), 3), 
             yb_sem=signif(sd(yb_meanC) / sqrt(length(N)), 3), 
             gr_mean=signif(mean(gr_meanC), 3), 
             gr_sem=signif(sd(gr_meanC) / sqrt(length(N)), 3), N=length(N))
)

(avg <- ggplot(data=typeGroup, aes(x=yb_mean, y=gr_mean, color=typeIDstr, shape=isPair)) +
  geom_point(size=3) + 
  geom_errorbar(aes(x=yb_mean, ymin=gr_mean - gr_sem, ymax=gr_mean + gr_sem)) +
  geom_errorbarh(aes(y=gr_mean, xmin=yb_mean - yb_sem, xmax=yb_mean + yb_sem, height=0.01)) + 
  coord_equal(xlim=c(-0.25, 0.25), ylim=c(-0.25, 0.25)) +
  xlab('') + 
  ylab('') +
  scale_color_manual(
    values = c(gray, gray, green, red, green, yellow, red), 
    name="spectral type",
    labels = c("unknown-cone", "unknown-pair", "M-cone", "L-cone", "M+M-pair", "L+M-pair", "L+L-pair")) +
  theme_bw(base_size=15))
ggsave("figures/lms_UAD_means.pdf")

```

Next, data was further grouped by the cone or pair targeted (masterID). The mean of each location is represented below.

```{r, fig.height=3, fig.width=11, message=FALSE}
singleAndPairs <- trial_data %>%
  select(., -typeIDstr) %>%
  group_by(., subject, masterID1, masterID2, isPair) %>% 
  summarise_all(., mean)

(ind <- ggplot(data=singleAndPairs, aes(x=yb, y=gr, color=as.factor(typeID), shape=isPair)) + 
  geom_point(alpha=0.76, size=0.8) +
  facet_grid(. ~ subject) +
  geom_rug(alpha=0.5) + 
  coord_equal(xlim=c(-1, 1), ylim=c(-1, 1)) + 
  xlab('(yellow - blue) / total') + ylab('(green - red) / total') + 
  scale_color_manual(
    values = c(gray, gray, green, red, green, yellow, red), 
    name="spectral type",
    labels = c("unknown-cone", "unknown-pair", "M-cone", "L-cone", "M+M-pair", "L+M-pair", "L+L-pair")) +
  #scale_color_brewer(palette = "Spectral") +
  geom_path(data = boundary, aes(x=x, y=y), colour='gray', linetype=2, inherit.aes = F) +
  theme_bw(base_size = 15) + 
  theme(legend.position="none"))

ggsave("figures/lms_UAD.pdf")

p <- grid.arrange(ind, avg, widths=2.5:1)
ggsave("figures/UAD_combined.pdf", p)
```

These plots illustrate the spread of responses collected from each cone or pair tested in each subject. There are a few features to note. Firstly, there were individual differences in color responses. S20075 used blue more frequently than the two other subjects and S10001 did not report yellow on any trials. However, the general patterns are similar. Most of the variance was found along the green-red dimension and there were few points that fell in the blueish-red or greenish-yellow quadrants These patterns are similar to previous reports (Schmidt et al. 2018). In the two subjects with classified mosaics, we additionally see that L-cone targeted trials tended to fall along the red region of this space, while M-cones were more likely to fall on the green side of the y-axis.

## Mosaic parameters do not predict responses

Also similar to previous findings, color reports varied even between cones with the same photopigment (Schmidt et al. 2018). Some L-cones, for instance, elicited highly saturated red sensations, while a majority produce white or desaturated pink reports. We next asked whether the variability in sensations between cones with the same sensitivity could be explained by low level features of the mosaic. Specifically, can we predict whether an L-cone will produce a saturated red or a desaturated pink based on the surrounding cone types or the delivery error? And in the case of cone pairs, did the distance between the two cones influence color appearance? To answer these question, we grouped responses according to the number of cones targeted (one or two), the session in which trial occurred and the specific cone(s) that was targeted. The mean and count was then computed for each cone(s). Only those cones pairs which had at least four good trials were analyzed.

```{r}
minTrials <- 4
singleConesSummary <- trial_data %>% 
  select(., -typeIDstr) %>%
  filter(., isPair==FALSE) %>% 
  group_by(., subject, session, masterID1) %>%   
  summarise_all(., c("mean", "length")) %>%
  filter(., yb_length >= minTrials) %>%
  select(., -contains("_length")) %>%
  rename_(.dots=setNames(names(.), gsub("_mean", "", names(.))))

twoConesSummary <- trial_data %>%   
  select(., -typeIDstr) %>%
  filter(., isPair==TRUE) %>% 
  group_by(., subject, session, masterID1, masterID2) %>%
  summarise_all(., c("mean", "length")) %>%
  filter(., yb_length >= minTrials) %>%
  select(., -contains("_length")) %>%
  rename_(.dots=setNames(names(.), gsub("_mean", "", names(.))))
```

Below is a matrix of correlation plots for the single cone condition.

```{r,message=FALSE}

ptest <- function(mat, ...) {
  # stolen shamelessly from: http://www.sthda.com/english/wiki/visualize-correlation-matrix-using-correlogram
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat<- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}



corr_analysis <- function (data, predictors) {
  # scatterplotMatrix with car package
  scatterplotMatrix(data[ , predictors], 
                    diagonal=list(method ="histogram"),
                    ellipse=FALSE)
  
  M<-cor(data[ , predictors], use="complete.obs")
  head(round(M,2))
  cor.mtest <- ptest

  # matrix of the p-value of the correlation
  p.mat <- cor.mtest(data[ , predictors])
  head(p.mat[, 1:5])

  corrplot(M, 
           type="upper",
           order="hclust", 
           p.mat = p.mat,
           sig.level = 0.01)

}

predictors = c("yb", "gr", "saturation", "lConeNeighbors", "distance_to_Scone_1", "distance_to_Scone_2")
corr_analysis(singleConesSummary, predictors)

```

Two cone condition:

```{r, fig.height=7, fig.width=7,  message=FALSE}

predictors = c("yb", "gr", "saturation", "lConeNeighbors",
               "distance_btwn_cones_arcmin",  "distance_to_Scone_1", "distance_to_Scone_2")
corr_analysis(twoConesSummary, predictors)
```

None of the above correlations were statistically significant with the exception of the relationship between gr and yb. The local neighborhood surrounding a cone is typically thought to be an important factor in generating color sensations. However, we did not find a statistical relationship. The distance between cones may also be an important factor influencing appearance. Neither color nor saturation judgments were correlated with the distance between targeted cones (plotted here in pixels). Cone pairs were never separated by more than one cone, which may explain why we did not detect a relationship. Moreover, the subjects verbally reported that the flashes always appeared as a single uniformly colored dot. Over these very small spatial distances, the visual system appears to be yoking neuronal activities together. In the future, systematically varying the distance between stimulated pairs will be an informative exercise. At a certain critical distance, the points of light will be seen as two spatially distinct dots. It is less clear at what distance the points will be perceived as two distinct colors.

## Two cone responses are an average of individual reports

While features of the mosaic and physical stimulus did not predict color reports, we hypothesized that the sensations recorded from individual cones would be predictive of paired stimulation conditions. To address this question, we matched the mean response from each cone pairs with the the mean report from each cone tested individually.

```{r, warning=FALSE,  message=FALSE}
merged1 <- merge(twoConesSummary, 
           select(singleConesSummary, subject, 
                  session, masterID1, yb, gr, lConeNeighbors), 
           by=c("subject", "session", "masterID1")) %>% 
  rename(., yb12=yb.x, gr12=gr.x, yb1=yb.y, gr1=gr.y,
                  lConeNeighbors12=lConeNeighbors.x, 
                  lConeNeighbors1=lConeNeighbors.y)

sessionMerge <- merged1 %>% 
  merge(., select(singleConesSummary, subject, session,
                  masterID1, yb, gr, lConeNeighbors), 
        by.x=c("subject", "session", "masterID2"),
        by.y=c("subject", "session", "masterID1")) %>% 
  rename(., yb2=yb, gr2=gr, lConeNeighbors2=lConeNeighbors) %>%
  mutate(., saturation12=abs(yb12) + abs(gr12),
         saturation1=abs(yb1) + abs(gr1),
         saturation2=abs(yb2) + abs(gr2))
```

We then fit a linear model to the data. Behavioral reports from two cone stimulation were predicted by an average of the individual responses: $gr_{12} = (gr_1 + gr_2) / 2$. Below is a visualization of the measured responses against the predictions.

```{r}
sessionMerge['predictedYB'] = (sessionMerge$yb1 + sessionMerge$yb2) / 2
sessionMerge['predictedGR'] = (sessionMerge$gr1 + sessionMerge$gr2) / 2
sessionMerge['predictedSaturation'] = abs(sessionMerge$predictedYB) + abs(sessionMerge$predictedGR)

# Add a column w difference from prediction
sessionMerge <- mutate(sessionMerge, 
                       diff_from_prediction=saturation12 - predictedSaturation)

```

```{r, fig.height=4, fig.width=9, message=FALSE}
unity = data.frame(x=c(-1, 1), y=c(-1, 1))
gr_plot <- sessionMerge %>%
  ggplot(., aes(x=predictedGR, y=gr12)) + 
  geom_point() +
  geom_smooth(method=lm) +
  coord_equal(xlim=c(-1, 1), ylim=c(-1, 1)) +
  geom_path(data=unity, aes(x=x, y=y), colour='gray', linetype=1) +
  ggtitle('(green - red) / total') +
  geom_path(data=unity, aes(x=x, y=y), colour='gray', linetype=1) +
  xlab('predicted response') + ylab('') + 
  theme_classic(base_size=15)

unity = data.frame(x=c(-1, 1), y=c(-1, 1))
yb_plot <- sessionMerge %>%
  ggplot(., aes(x=predictedYB, y=yb12)) + 
  geom_point() +
  geom_smooth(method=lm) +
  coord_equal(xlim=c(-1, 1), ylim=c(-1, 1)) +
  geom_path(data=unity, aes(x=x, y=y), colour='gray', linetype=1) +
  ggtitle('(yellow - blue) / total') +
  geom_path(data=unity, aes(x=x, y=y), colour='gray', linetype=1) +
  xlab('predicted response') + 
  ylab('observed pair response') + 
  theme_classic(base_size=15)

p <- grid.arrange(yb_plot, gr_plot, nrow=1)

ggsave("figures/prediction.pdf", p)

```

Results of linear regression (gr dimension followed by yb):
```{r}
# A linear model predicts the results based on an average of the individual responses.
mod <- lm(gr12 ~ predictedGR, sessionMerge)
summary.lm(mod)

mod <- lm(yb12 ~ predictedYB, sessionMerge)
summary.lm(mod)
```


## Pairs with the same type are more saturated than predicted by average

A simple linear model captured a large fraction of the variance (>72%). However, there were some pairs that deviated substantially from the best fit line. We wondered whether the deviation from linearity might be predicted by the sub-class of the two cones. For instance, do an L- and M-cone interact in a non-linear manner, while two L or two M-cones sum linearly?

We found the saturation for each pair and subtracted it from the average of the two cones probed alone. Those results are plotted below. A unity line represents the condition where the observed saturation judgment was predicted exactly by the average of individual responses. Notice that the L+L and M+M pairs tend to fall above the unity line -- particularly at higher saturation values. In contrast, the L+M pairs fall below the line. These observations indicate that the cones with the same spectral type produced slightly more saturated reports than predicted by the average of their individual responses.

```{r, fig.height=3, fig.width=5, message=FALSE}

(scatLM <- sessionMerge %>%
   filter(., typeID == 1 | typeID > 3) %>%
   ggplot(., 
       aes(x=predictedSaturation, y=saturation12, color=as.factor(typeID))) + 
  geom_point() + 
  coord_equal(xlim=c(0, 1), ylim=c(0, 1)) +
  scale_color_manual(values = c(gray, green, yellow, red), 
                     name="spectral type",
                     labels = c("unknown", "M+M", "L+M", "L+L")) +
  geom_path(data=unity, aes(x=x, y=y), colour='gray', linetype=1) + 
  xlab('predicted saturation') +
  ylab('measured saturation') +
  theme_classic(base_size=15) +
  theme(legend.position='none')
)

ggsave('figures/observed-predicted_scatter.pdf')

```
```{r, fig.height=3, fig.width=5, message=FALSE}

scatUn <- sessionMerge %>%
   filter(., typeID == 1) %>%
   ggplot(., aes(x=predictedSaturation, y=saturation12)) +  
   geom_point(color=gray) +  
   coord_equal(xlim=c(0, 1), ylim=c(0, 1)) +
   #scale_color_manual(values = c("#3aaf4f", "#fff247", "#bf3535"), name="spectral type",
   #                   labels = c("M+M", "L+M", "L+L")) +
   geom_path(data=unity, aes(x=x, y=y), colour='gray', linetype=1) + 
   xlab('predicted saturation') +
   ylab('measured saturation') +
   theme_classic(base_size=15)

ggsave('figures/observed-predicted_scatter_unclassified.pdf')

```

We quantified this trend directly by taking the difference between the observed and predicted saturation judgments. The results are illustrated in a histogram below. Student's t-test's confirm that the L+M pairs were significantly less saturated (more white) than a simple average of their individual responses, while the opposite was true for cones with the same spectral type.

```{r, fig.height=3, fig.width=5, message=FALSE}
diffHist <- sessionMerge %>%
   filter(., typeID == 1 | typeID > 3) %>%
   ggplot(., 
          aes(x=saturation12 - predictedSaturation, 
              color=as.factor(typeID))) + 
  geom_freqpoly(binwidth=0.1, size=1.25) +
  scale_color_manual(
    values = c(gray, green, yellow, red), 
    name="spectral type",
    labels = c("unknown-pair", "M+M", "L+M", "L+L")) + 
  xlab("measured - predicted saturation") +
  theme_classic(base_size=15) +
  theme(legend.position=c(0.75, 0.80))

ggsave('figures/observed-predicted_histogram.pdf')
```

```{r,fig.height=4, fig.width=9, echo=FALSE, message=FALSE}
(p <- grid.arrange(scatLM, diffHist, nrow=1))

ggsave('figures/observed-predicted.pdf', p)

```

L+M pairs have a mean that is significantly different from zero.

```{r}

t.test(sessionMerge %>% 
         filter(., typeID == 5) %>% 
         select(., diff_from_prediction)
       )
```

L+L and M+M pairs have a mean that is significantly different from zero, but in the opposite direction.

```{r}
t.test(sessionMerge %>% 
         filter(., typeID == 4 | typeID == 6) %>%
         select(., diff_from_prediction)
       )
```

```{r}

t.test(sessionMerge %>% 
         filter(., typeID == 1) %>% 
         select(., diff_from_prediction)
       )
```

Finally, L+M pairs are significantly different from the population of responses collected from L+L and M+M pairs.
```{r}
t.test(sessionMerge %>% 
         filter(., typeID == 5) %>% 
         select(., diff_from_prediction),
       sessionMerge %>% 
         filter(., typeID == 4 | typeID == 6) %>%
         select(., diff_from_prediction)
       )
```

# Conclusions

Color sensations from pairs of cones were predicted by a simple average of individual responses. However, when two cones from the same subclass were probed, there was a systematic deviation from a simple average. These pairs produced significantly more saturated colors than predicted by an average of the colors elicited when probed alone. This observation suggests that the visual system uses a different strategy when combining information within versus across neuronal sub-classes.

